---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Vignette Info

The purpose of this vignette is to explain the withdrawal/dropout modelling assumptions used in the package.

## Assumptions

Suppose that survival time $T$ follows a piecewise exponential model with rates $(\lambda_1, \lambda_2,\ldots, \lambda_k$) and changepoints $(t_1,\ldots,t_{k-1})$.

Suppose that dropout time $C$ is independent and follows an exponential distribution with rate $\mu$.

## What are we trying to calculate?

We are interested in the probability that no event is observed by time $t$.


$$pr\left( \left\lbrace T >t \text{ and }C > t\right\rbrace \text{or} \left\lbrace  T > C \text{ and }C < t\right\rbrace \right) = pr\left( T >t \text{ and }C > t\right) + pr\left( T > C \text{ and }C < t \right)$$

The first term:

$$pr\left( T >t \text{ and }C > t\right) = \exp(-\sum \lambda_i g(t, t_i))\exp(-\mu t)$$


where $g(t, t_i) = \max(0, \min(t - t_{i - 1}, t_i - t_{i-1}))$.


The second term:

$$  pr\left( T > C \text{ and }C < t \right) = \int_0^t pr(T > c)f(c)dc$$


Consider the case with just one change point, first when $t < t_1$:

$$\int_0^t pr(T > c)f(c)dc ~~= \int_0^{t}\exp(-\lambda_1c)\mu\exp(-\mu c)dc$$
$$ = \left( \frac{\mu}{\mu + \lambda_1} \right)\left[1 - \exp\left\lbrace-(\mu + \lambda_1)t\right\rbrace\right]$$

Now, when $t > t_1$:

$$\int_0^t pr(T > c)f(c)dc ~~= \int_0^{t_1}\exp(-\lambda_1c)\mu\exp(-\mu c)dc + \int_{t_1}^t\exp\left\lbrace-\lambda_2(c - t_1) - \lambda_1t_1\right\rbrace\mu\exp(-\mu c)dc$$

The first term on the right-hand-side is:

$$ \left( \frac{\mu}{\mu + \lambda_1} \right)\left[1 - \exp\left\lbrace-(\mu + \lambda_1)t_1\right\rbrace\right].$$

The second term on the right-hand-side is equal to:

$$-\left( \frac{\mu}{\mu + \lambda_2} \right) \left[ \exp\left\lbrace -\lambda_2(t - t_1)-\lambda_1t_1-\mu t\right\rbrace  - \exp\left\lbrace-(\mu + \lambda_1)t_1 \right\rbrace \right] $$


$$= \left( \frac{\mu}{\mu + \lambda_2} \right) \exp\left\lbrace-(\mu + \lambda_1)t_1\right\rbrace\left[1 - \exp\left\lbrace-(\mu + \lambda_2)(t - t_1)\right\rbrace\right]$$

And it's easy to see how this formula is extended for more changepoints. This formula makes sense: we calculate the probability that $\min(T,C)$ occurs in an interval $t_{i-1}, t_i$, multiply this by the probability the minimum was $C$ rather the $T$ (which is just $\mu / (\mu + \lambda_i)$), and then we sum over the intervals up to time $t$.


## Computation


```{r}
devtools::load_all("..")
surv_pieces(t = 5, 
            change_points = 10,
            lambdas = c(log(2) / 15, 
                        log(2) / 15), 
            mu = 0.12, 
            dropouts = FALSE)
```


## What if we are interested in the number of dropouts?

If, instead, we were interested in the probability that a patient has not dropped out by time $t$ we would simply switch 

$$\frac{\mu}{\lambda_i + \mu}$$

for 

$$\frac{\lambda_i}{\lambda_i + \mu}$$

in the above formulae. We can do this with the `surv_pieces` function by change `dropouts` to `TRUE`:

```{r}
surv_pieces(t = 5, 
            change_points = 10,
            lambdas = c(log(2) / 15, 
                        log(2) / 15), 
            mu = 0.12, 
            dropouts = TRUE)
```


