---
title: "expectedevents: tutorial"
author: "Dominic Magirr"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## What problem does this solve?

When designing a trial with a time-to-event endpoint, we often want to know how many events are expected at a given calendar time.

We will typically have some information on recruitment.

```{r}
#### recruitment information
recruitment_1 = list(n = 100, 
                     r_period = 12, 
                     k = 1)
```

In this case, a single arm trial with 100 patients recruited over 12 months. The parameter $k$ dictates the shape of the recruitment curve:

$$pr(\text{recruited by calendar time t}) = (t / \text{r_period}) ^ k.$$

We also usually have a working assumption for the survival model:


```{r}
#### model for survival distribution
model_1 = list(change_points = 4, 
               lambdas = c(log(2) / 15, log(2) / 15))

```


In this package, we only allow piece-wise exponential models with at least 2 pieces. This is not really a restriction because we can easily approximate any parametric survival distribution. In the above example, I'm assuming an exponential model with median 15 months. To do this, I specify an arbitrary change-point, but the event rate either side of the change-point remains the same.


How many events at a data cut-off (`dco`) at 30 months?

```{r}
#library(expectedevents)
devtools::load_all("..")
expected_events_single_arm(dco = 30,
                           recruitment = recruitment_1,
                           model = model_1)
```


## Two-arm trial

We can do the same thing for a two-arm trial

```{r}
recruitment_2 = list(n_0 = 100, 
                     n_1 = 100, 
                     r_period = 12, 
                     k = 1)


model_2 = list(change_points = 4, 
               lambdas_0 = c(log(2) / 15, log(2) / 15), 
               lambdas_1 = c(log(2) / 15, 0.7 * log(2) / 15))


```

In this case the hazard ratio is 1 for the first 4 months. Thereafter, the hazard ratio is 0.7.

```{r}
expected_events_two_arm(dco = 30,
                        recruitment = recruitment_2,
                        model = model_2)

```


## Dropouts

We can add a parameter to deal with patients who dropout (withdraw from the study and are lost to follow-up). We assume that dropout occurs independently to the event of interest, and that time-to-dropout follows an exponential distribution with rate `mu`.

For example, we can modify the above example by adding dropout at rate `mu = 0.01`:

```{r}
expected_events_two_arm(dco = 30,
                        recruitment = recruitment_2,
                        model = model_2,
                        mu = 0.01)

```

We observe fewer events due to the competing risk of dropout. If we acutally want to see the expected number of dropouts, we need to add the argument `dropouts = TRUE`:

```{r}
expected_events_two_arm(dco = 30,
                        recruitment = recruitment_2,
                        model = model_2,
                        mu = 0.01,
                        dropouts = TRUE)

```

## Proportion of patients who have withdrawn

The dropout parameter `mu` does not have a very intuitive interpretation. Often, we would prefer to assume something along the lines of: "we expect 10% of patients (out of the final total sample size) to have withdrawn at a data cut-off of 30 months".

`prop_to_mu_single_arm` and `prop_to_mu_two_arm` are functions that can convert this statement into a value for `mu`.

```{r}
prop_to_mu_two_arm(prop_withdrawn = 0.1,
                   dco = 30,
                   recruitment = recruitment_2,
                   model = model_2)
```

## Multiple data-cut offs

The `expected_events_single_arm` and `expected_events_two_arm` functions only allow a single `dco` as first argument. To assess several data cut-offs I suggest using, e.g., `purrr::map`:


```{r}
dcos = seq(0.001, 30, length.out = 10)

events_dco = purrr::map(dcos,
                        expected_events_single_arm,
                        recruitment = recruitment_1,
                        model = model_1,
                        mu = 0.01)

dropouts_dco = purrr::map(dcos,
                          expected_events_single_arm,
                          recruitment = recruitment_1,
                          model = model_1,
                          mu = 0.01,
                          dropouts = TRUE)

round(unlist(events_dco), 1)
round(unlist(dropouts_dco), 1)
```

## Average hazard ratio

For 2-arm trials, the `expected_events_two_arm` function also calculates the 'average hazard ratio', which is defined as

$$\exp\left\lbrace \sum_i p_i \log \left(  \frac{\lambda_{1,i}}{\lambda_{0,i}} \right) \right\rbrace$$

where $\lambda_{0,i}$, $\lambda_{1,i}$ are the event rates on the two arms during patient-time interval [$t_i, t_{i-1}$), and $p_i$ is the proportion of total events that occur during patient-time interval [$t_i, t_{i-1}$). Here, $t_1,\ldots,t_{k-1}$ correspond to `change_points`. Also, $t_0 = 0$ and $t_{k} = \infty$.


To see this additional information we need to add the argument `total_only = FALSE`:

```{r}
expected_events_two_arm(dco = 30,
                        recruitment = recruitment_2,
                        model = model_2,
                        mu = 0.01,
                        total_only = FALSE)

```

## Approximating a survival curve from a different model

Let's take the example of a cure rate model. Consider an adjuvant study where 70% of patients are already "cured" coming in. Suppose for the "non-cured" patients, the hazard rate is $\log(2)/12$ on the control arm and $\log(2) / 6$ on the experimental arm. Overall, the hazard rates are non-proportional...

```{r}
# survival
s_c = function(t) 0.7 + 0.3 * exp(-log(2) / 12 * t)
s_e = function(t) 0.7 + 0.3 * exp(-log(2) / 6 * t)
# density
f_c = function(t) log(2) / 12 * 0.3 * exp(-log(2) / 12 * t)
f_e = function(t) log(2) / 6 * 0.3 * exp(-log(2) / 6 * t)
# hazard
h_c = function(t) f_c(t) / s_c(t)
h_e = function(t) f_e(t) / s_e(t)
# hazard ratio
hr = function(t) h_e(t) / h_c(t)

ts = seq(0, 48, by = 1)

plot(ts, h_c(ts),
     type = 'l',
     ylim = c(0,0.05),
     ylab = "hazard",
     xlab = "time")

points(ts,
       h_e(ts),
       type = 'l',
       col = 2)

legend("topright",
       c("control", "experimental"),
       lty = c(1,1),
       col = 1:2)

### approximate model

model_approx = list(change_points = seq(2, 46, by = 2), 
                    lambdas_0 = h_c(seq(1, 47, by = 2)), 
                    lambdas_1 = h_e(seq(1, 47, by = 2)))

points(c(0, model_approx$change_points), model_approx$lambdas_0, type = 's')
points(c(0, model_approx$change_points), model_approx$lambdas_1, type = 's', col = 2)


```



We can use this approximate model, as before.

```{r}
expected_events_two_arm(dco = 30,
                        recruitment = recruitment_2,
                        model = model_approx,
                        mu = 0.01)

```

## Example how to plot survival curves from approximation

```{r}
plot(seq(0, 48, length.out = 100),
     unlist(purrr::map(seq(0, 48, length.out = 100), 
                       surv_pieces,
                       change_points = model_approx$change_points,
                       lambdas = model_approx$lambdas_0)),
     type = 'l',
     ylim = c(0,1),
     ylab = "survival",
     xlab = "time")
```


## Calculating power for a GSD under NPH

```{r}
library(gsDesign)
library(mvtnorm)
## get timing
event_targets = c(60, 120)

dcos = purrr::map_dbl(event_targets, expected_dco_two_arm, 
                      recruitment = recruitment_2,
                      model = model_2,
                      mu = 0,
                      dropouts = FALSE)

## get average hazard ratios
full_info = purrr::map(dcos, expected_events_two_arm,
                       recruitment = recruitment_2,
                       model = model_2,
                       mu = 0,
                       dropouts = FALSE,
                       total_only = FALSE)

av_hrs = purrr::map_dbl(full_info, function(x) x$av_hr)

## get bound
d = gsDesign(k = 2, test.type = 1, sfu = sfLDOF)
d$upper$bound

p = get_power(events = event_targets,
              av_hrs = av_hrs,
              upper_bound = c(d$upper$bound),
              lower_bound = c(0, c(d$upper$bound)[2]),
              recruitment = recruitment_2)

p$power_so_far
cumsum(p$upper_stopping)

p$upper_stopping
p$lower_stopping

```







 
